Fast Fourier transformation of electromagnetic data for computationally expensive kernels

### 2021-04-07 ManuscriptTime.tex
Revision II - clean

\documentclass[extra, camera,%
    onecolumn,   % comment for two columns
    referee,     % uncomment for double lines
    % final,       % uncomment for changes
]{gji}
\usepackage{timet}

@@

\begin{summary}
%
Three-dimensional controlled-source electromagnetic data is often computed
directly in the domain of interest, either in the frequency domain or in the
time domain. Computing it in one domain and transforming it via a Fourier
transform to the other domain is a viable alternative. It requires the
evaluation of many responses in the computational domain if standard Fourier
transforms are used. This can make it prohibitively expensive if the kernel is
time-consuming as is the case in three-dimensional electromagnetic modelling.
The speed of such modelling that is obtained through a transform is defined by
three key points: solver, method and implementation of the Fourier
transform\deleted{ation}, and gridding.
%
The faster the solver, the faster modelling will be. It is important that the
solver is robust over a wide range of values (frequencies or times). The method
should require as few kernel evaluations as possible while remaining robust. As
the frequency and time ranges span many orders of magnitude, the required
values are ideally equally spaced on a logarithmic scale. The proposed fast
method uses either the digital linear filter method or the logarithmic fast
Fourier transform together with a careful selection of evaluation points and
interpolation. In frequency-to-time domain tests this methodology requires
typically 15 to 20 frequencies to cover a wide range of offsets.
%
The gridding should be frequency- or time-dependent, which is accomplished by
making it a function of skin depth. Optimising for the least number of required
cells should be combined with optimising for computational speed. Looking
carefully at these points results in much smaller computation times with
speedup factors of ten or more over previous methods. A computation in one
domain followed by transformation can therefore be an alternative to
computation in the other domain domain if the required evaluation points and
the corresponding grids are carefully chosen.
%
\end{summary}

@@

The controlled-source electromagnetic (CSEM) method is one of the common
non-seismic tools in exploration geophysics, not only in hydrocarbon
exploration \citep{GEO.10.Constable}, but also in the search for sulfides
\citep{GRL.19.Gehrmann}, water \citep{GEO.05.Pedersen}, geothermal sources
\citep{WGC.15.Girard}, or for geological purposes \citep{NAT.19.Johanson}.
While current sources with a few frequencies are used in the deep marine
environment, transient measurements are more common in the shallow marine
environment and on land \citep[e.g., ][]{GEO.07.Ziolkowski, SEG.07.Andreis,
SEG.07.Avdeeva}. One of the main reasons is the dominance of the airwave in
shallow marine and terrestrial measurements, which can be better separated in
the time domain. CSEM is usually divided into frequency- and time-domain
methods, depending on whether the source signal is a continuous waveform, such
as a sine, or a finite waveform, such as a pseudo-random binary sequence
(PRBS). A numerical comparison of the two methods is given by
\cite{GP.13.Conell}. Acquired CSEM data are subsequently often analysed
(modelled and inverted for) in their respective domain, either the frequency or
time domain. Modellers of layered media usually exploit the horizontal
shift-invariance by computing the responses in the wavenumber-frequency domain
followed by a 2D inverse spatial Fourier transform, also called Hankel
transform, to the space-frequency domain, and a regular inverse Fourier
transform if time-domain data are required \citep[e.g., ][]{GEO.15.Hunziker}.
CSEM codes for 2D and 3D computations, on the other hand,
\replaced{often}{mostly} compute their responses directly in the required
domain, either frequency or time.

\added{Electromagnetic methods in geophysics span a wide range of model
scenarios and acquisition layouts, each with its own modelling-related
implications. A well-known example is the aforementioned land or shallow marine
case versus the deep marine case, where in the former case a dominating airwave
has to be accurately modelled, whereas it can be completely ignored in the
latter case. Recent areas of particular interest in time-domain modelling that
pose numerical challenges are, for instance, simulating the fields through
steel-cased wells \mbox{\citep{CAG.19.Heagy}} or the effects of induced
polarization \mbox{\citep{EXP.20.Kang}}. The former is challenging because of
the high conductivity contrasts requiring very detailed meshing, the latter is
challenging for time-domain modelling because the models are frequency
dependent.}

\deleted{There is a wealth of 3D electromagnetic codes, with a first boom in the
1990's; \mbox{\cite{B.SEG.99.Oristaglio}} provide a good review of the state of
art at that time. A second wave occurred during the CSEM boom in the
hydrocarbon industry in the early 21th century, see, e.g.,
\mbox{\cite{SG.05.Avdeev}} and \mbox{\cite{SG.10.Borner}} for comprehensive
overviews. Recent publications after these review articles span the width of
numerical methods from integral equation codes \mbox{\citep{MGS.17.Kruglyakov}}
and differential equation methods such as finite differences
\mbox{\citep{GEO.10.Mittet, GEO.11.Zaslavsky, CAG.13.Sommer}}, finite elements
\mbox{\citep{GJI.13.Grayver}}, and finite volumes
\mbox{\citep{GEO.14.Jahandari}}. Besides the different methods there are also
many different types of discretisation. Commonly used, besides rectilinear Yee
grids \mbox{\citep{IEEE.66.Yee}} as used by our code, are Lebedev grids
\mbox{\citep{CMMP.64.Lebedev}} but also unstructured grids using, for instance,
tetrahedra \mbox{\citep{CAG.17.Cai}}. The computational requirement to model
CSEM data also depends heavily on the technique used to solve the system. There
are direct solvers \mbox{\citep{GEO.15.Grayver}}, indirect solvers
\mbox{\citep{GJI.15.Jaysaval}}, and more recently also hybrid solvers
\mbox{\citep{GEO.18.Liu}}.}

Under certain conditions it can be a viable alternative to model time-domain
data with a frequency-domain code, as shown by \cite{GEO.08.Mulder}. They
compared the computational complexity \added{of different time-domain methods
with the computational complexity of frequency-domain computation followed by a
Fourier transform:} \deleted{of} explicit time-stepping schemes (Du
Fort-Frankel method as used in, e.g., \cite{GEO.04.Commer,
GEO.07.Maao})\replaced{ and}{;} implicit schemes (e.g.,
\cite{GEO.04.Haber}\added{; \mbox{\cite{GEO.10.Um}}}) \added{ have a complexity
of $O(n^4)$}, \deleted{and} matrix exponentials and Lanczos reduction schemes
\cite[e.g., ][]{RS.94.Druskin} \added{have a complexity of $O(n^4\sqrt{\log
n})$}\replaced{, and}{with the} Fourier transform of frequency-domain solutions
\added{have a complexity $O(n_f n^3)$}; \added{$n$ is the number of grid points
in each direction, and $n_f$ the number of frequencies}. Their conclusion was
that the Fourier transform method can be favourable if the number of required
frequencies is small relative to the number of grid points in each coordinate.
The conditions for fast computation of time-domain data with a frequency-domain
code are: a sufficiently powerful solver, appropriate frequency selection and
interpolation, and an automated gridding, for which they used the
multi-frequency CSEM approach presented by \cite{GEO.07.Plessix}. We build upon
these results but improve the run time from hours to minutes. The main reasons
for this significant speed-up are \added{the further reduction of required
frequencies by introducing lower and upper thresholds of numerically important
frequencies,} an adaptive, frequency-dependent gridding scheme that minimizes
the required cells in each dimension, and a logarithmic Fourier transform such
as digital linear filters \citep[DLF, ][]{GP.71.Ghosh} or the logarithmic fast
Fourier transform \citep[FFTLog, ][]{RAS.00.Hamilton} to go from the frequency
to the time domain. The latter makes it also possible to only use the imaginary
part of the frequency-domain response, which has advantages when it comes to
interpolation.

\deleted{While we use the finite-integration technique
\mbox{\citep{AEU.77.Weiland}} in the frequency domain to demonstrate our
method, it is important to note that our ideas can be used with any solver and
is not bound to one or another frequency- or time-domain solver. The ideas for
the frequency selection can be applied to time selection, and time-dependent,
adaptive gridding exists as well \mbox{\citep[e.g., ][]{RS.06.Commer}.}} In the
next section, we briefly review the methodology as introduced by
\cite{GEO.07.Plessix} and \cite{GEO.08.Mulder} and highlight their advantages
and shortcomings. This is followed by an outline of \added{the methodology,}
our changes to the Fourier transform and the adaptive gridding. Finally, we
demonstrate the efficiency of the approach with some numerical results.
\added{These compromise of a homogeneous space, a layered model, a model which
includes dispersive media (induced polarization), and a three-dimensional
model.}

@@

Being able to model CSEM data both in the frequency domain and in the time
domain can be desirable, as both domains have advantages and disadvantages. One
way to achieve this is to implement Maxwell's equations in both domains
\added{as is done, for instance, in \texttt{SimPEG}
\mbox{\citep{CAG.15.Cockett}} or in \texttt{custEM}
\mbox{\citep{GJI.21.Rochlitz}}}. Another approach is to have Maxwell's equations
only implemented in one domain, and use Fourier transforms to go to the other.
However, this approach can be costly, as many frequencies over a wide range are
required to go from the frequency domain to the time domain, or many times over
a wide range for the opposite direction. We present a methodology which
significantly reduces the amount and range of the required
frequencies\replaced{ through a combination of extrapolation, interpolation,
and setting the responses for high frequencies to zero. This translates}{,
translating} into a significant reduction in computation time. \replaced{The
number of required frequencies is not the only important point for reducing
computation time. Another important aspect is the computation grid, as t}{T}he
required \deleted{computation} grids for low frequencies (in our case in the
order of 0.001\,Hz) and high frequencies (around 100\,Hz) are hugely different.
Low frequencies can be computed on a coarser grid, but they require a much
larger domain in order to avoid boundary effects. High frequencies, on the
other hand, require denser gridding, but they are much more limited in reach.
An adaptive gridding scheme is therefore required\replaced{, which}{. We build
our approach upon \mbox{\cite{GEO.07.Plessix}}, who presented such an adaptive
gridding for multi-frequency (and multi-source) CSEM modelling.

An adaptive grid in the frequency domain} is naturally based on the skin depth,
\deleted{which is} the distance after which the amplitude of the
electromagnetic field \replaced{has}{is} decayed by $1/e\approx 37\,\%$. The
skin depth $\delta$ is a function of conductivity and frequency, and for the
diffusive approximation of Maxwell's equations in an isotropic, homogeneous
medium is given by \citep[e.g.,][\mbox{\deleted{
equation~1.53}}]{B.SEG.88.Ward}
%
\begin{equation}
  \delta = \sqrt{\frac{2}{\omega\mu\sigma}}
  \ \approx \ 503.3/\sqrt{f\sigma} \, ,
  \label{eq:skindepth}
\end{equation}
%
where $\sigma$ is conductivity (S/m), $\omega=2\pi f$ is angular frequency of
frequency $f$ (Hz), and $\mu$ is magnetic permeability (H/m). The approximation
is obtained by using the free-space value of magnetic permeability,
$\mu_0=4\pi\times10^{-7}\,$H/m.

\added{We build our approach upon} \cite{GEO.07.Plessix}\added{, who presented
such an adaptive gridding for multi-frequency (and multi-source) CSEM
modelling. They} define the minimum cell width $\Delta_\mathrm{min}$ as
a fraction of the minimum skin depth $\delta_\mathrm{min}$, where
$\delta_\mathrm{min}$ should be 2--3 times bigger than $\Delta_\mathrm{min}$.
The cells have to be smallest around the source; in the marine case, the
minimum skin depth is therefore defined by the conductivity of seawater.
However, this \replaced{can yield}{yields} quite large cells for low
frequencies, so special care has to be taken around the source by defining a
maximum allowed $\Delta_\mathrm{min}$. The grid dimension, on the other hand,
is defined as a function of \deleted{the} skin depth for the average
conductivity of the background, $\delta_\mathrm{ave}$. They use four times
$\delta_\mathrm{ave}$ for the $x$-, $y$-, and downward $z$-directions, and a
fixed 50\,km for the upward $z$-direction to account for the airwave. To reduce
the number of cells, it is desirable to introduce stretching, at least in the
buffer zone outside of the area where source and receivers are located.

@@

We present improvements to both the adaptive gridding and the transform from
frequency domain to time domain \added{using the finite-integration technique
\mbox{\citep{AEU.77.Weiland}}}, which makes time-domain modelling with a
frequency-domain code viable in comparison with time-domain codes.
It is important to note that \added{our ideas can be used with any solver and
is not bound to one or another frequency- or time-domain solver. The ideas for
the frequency selection can be applied to time selection, and time-dependent,
adaptive gridding exists as well \mbox{\citep[e.g., ][]{RS.06.Commer}.}}
\deleted{while we show the transform from frequency to time domain, the
conclusions regarding the Fourier transform can equally well be applied to the
transform from time-domain to frequency-domain data for a code that solves
Maxwell's equations in the time domain.} \added{Our recommendations for
speeding up the Fourier transform for expensive kernels are independent of
spatial complexity, as diffusive EM responses are smooth functions of both
frequency and time. However, the intrinsic requirement is an accurate solver
over a sufficient wide range of frequencies or times.} \deleted{A variation of
our method was implemented, for instance, by \custem
\mbox{\citep{GEO.19.Rochlitz}} to obtain frequency-domain data from time-domain
computations (Rochlitz, private communication, 2020).}

@@

\added{The requirement for any transform is a robust solver of sufficient
accuracy over a wide range of frequencies.} \replaced{We use f}{F}or the
numerical computations \deleted{we use} the open-source (Apache License 2.0)
code\added{s \empymod} \added{\mbox{\citep{GEO.17.Werthmuller}}} \added{and}
\emg3d \citep{JOSS.19.Werthmuller}\replaced{.}{,} \added{The former computes
semi-analytical responses of layered models, the latter is} a \added{3D}
multigrid solver based on \cite{GP.06.Mulder}, which can be used as a
preconditioner for Krylov subspace solvers or as a solver on its own. The
multigrid approach works fine for the diffusive approximation of Maxwell's
equations, which assumes that $\omega\varepsilon \ll \sigma$, where
$\varepsilon$ is electric permittivity (F/m). The remaining system to solve in
the frequency domain is then given by the second-order differential equation of
the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which known improvements such as line-relaxation and semicoarsening
\citep{ECCFD.06.Jonsthovel} are implemented, with a non-standard Cholesky
decomposition to speed up the computation \citep{GEO.08.Mulder}. One of the big
advantages of the multigrid method is that it scales linearly (optimal) with
the grid size in both CPU and RAM usage \citep{B.Springer.20.Mulder}. This
makes it feasible to run even big models on standard computers, without the
need for big clusters. All \deleted{the} examples in this article are run on a
laptop with an i7-6600U\,CPU\,@\,2.6\,GHz (x4) and 16\,GB of memory, using
Ubuntu 20.04 and Python 3.8. \replaced{They}{All examples} were run using a
single thread, but parallelization over frequencies is straight forward, as the
computations are independent. \deleted{For the semi-analytical computations of
layered models we use the open-source (Apache License 2.0) code \empymod
\mbox{\citep{GEO.17.Werthmuller}}.}

\added{Note that while we use these two algorithms for the numerical examples,
any solver that provides sufficiently accurate frequency-domain responses could
be used.}

@@

An important factor in terms of speed and accuracy for time-domain responses
obtained from frequency-domain computations is the selection of the required
frequencies. The fewer frequencies required, the quicker we obtain the
time-domain result. \added{This is true in general but applies in particular
to expensive kernels such as is the case in 3D modelling, where the computation
of the response of a single frequency takes much longer than the transformation
itself.} The same applies for the reverse operation, the fewer times required,
the quicker we obtain frequency-domain results. Even though we limit our
analysis here to frequency-to-time transformations, most of the arguments are
reversible and applicable to time-to-frequency transformations. We decided to
use a regular spacing of frequencies on a log-scale, rather than an adaptive
scheme. This approach is favourable if a wide range of offsets is needed, as
the required frequencies change with offset and an adaptive frequency selection
is therefore often tailored to a single offset\added{ or a limited range of
offsets}. Also, it allows for straightforward parallelization over frequencies,
which is not completely possible with an adaptive scheme.

@@

\replaced{Any Fourier transform needs a certain range of frequencies to obtain
time-domain results. Our approach is not to minimize the required frequencies
for the transform, but to minimize the number of frequencies for which we
actually have to compute responses. We do this in three zones: above an upper
threshold $f_\mr{max}$, we set the frequency-domain response to zero. Below a
lower threshold $f_\mr{min}$, we extrapolate the frequency-domain response. As
we know that the imaginary response goes towards zero for zero frequency, we
can turn it into an interpolation, for which we use PCHIP. And in-between the
thresholds, we use cubic spline interpolation in order to only have to actually
compute a few distinct frequencies per decade ($n_\mr{dec}^f$). Using this
approach, we found that computing the responses for 15--25 frequencies is
usually good enough to obtain time-domain responses for a large range of
offsets; even though the actual transform might need up to hundreds of
frequencies.}{There are three key-parameters that have to be defined: Minimum
and maximum frequency ($f_\mr{min}, f_\mr{max}$) and the number of frequencies
per decade ($n_\mr{dec}^f$).}

\replaced{The required frequencies depend naturally on the conductivities and
on the chosen acquisition type and layout. A t}{T}rial-and-error approach with a
3D code is very time-intensive. However, a simplified, layered model for the
required survey setup and a fast 1D modeller makes it possible to estimate
these parameters easily. \replaced{An example of this is shown in
Figure~\ref{fig:FFTLog-log}.}{To this end, we created a graphical, interactive
tool using \empymod, as shown in Figure~\ref{fig:FFTLog-log}.}
%
\begin{figure}
  \centering
  \includegraphics[width=\cwidth]{01-FFTLog-log}
  \caption{\replaced{Example}{Interactive} frequency selection for a
    \replaced{simple}{user-provided} layered model (the shown model parameters
    are described in the text). This example shows the impulse response at an
    offset of 5\,km, for which it uses FFTLog from 0.001\,Hz to 10\,Hz with
    five frequencies per decade. \deleted{Note that the error is clipped for
    values smaller than 0.01\,\% and bigger than 100\,\%.}}
  \label{fig:FFTLog-log}
\end{figure}
%
\replaced{It shows the responses for}{The shown example model is} a
marine scenario with 1\,km water depth of resistivity $\rho=0.3\,\ohmm$ ($\rho
= \sigma^{-1}$) and a 100\,m thick target of 100\,\ohmm at 1\,km below the
seafloor in a background of 1\,\ohmm. The source is 50\,m above the seafloor,
the receiver is on the seafloor, and the response is the inline $x$-directed
$E$-field. The left subplot shows the imaginary part in the frequency domain
and the right plot the corresponding impulse response in the time domain. The
red lines are the semi-analytical responses. The blue circles indicate the
actually computed responses and the black dots the frequencies which are
interpolated or set to zero. The resulting time-domain response has a relative
error of less than 1\,\% everywhere except for very early times.

>>>>

\documentclass[extra, camera,%
    % onecolumn,   % comment for two columns
    % referee,     % uncomment for double lines
    final,       % uncomment for changes
]{gji}
\usepackage{timet}

@@

\begin{summary}
%
Three-dimensional controlled-source electromagnetic data is often computed
directly in the domain of interest, either in the frequency domain or in the
time domain. Computing it in one domain and transforming it via a Fourier
transform to the other domain is a viable alternative. It requires the
evaluation of many responses in the computational domain if standard Fourier
transforms are used. This can make it prohibitively expensive if the kernel is
time-consuming as is the case in three-dimensional electromagnetic modelling.
The speed of such modelling that is obtained through a transform is defined by
three key points: solver, method and implementation of the Fourier transform,
and gridding.
%
The faster the solver, the faster modelling will be. It is important that the
solver is robust over a wide range of values (frequencies or times). The method
should require as few kernel evaluations as possible while remaining robust. As
the frequency and time ranges span many orders of magnitude, the required
values are ideally equally spaced on a logarithmic scale. The proposed fast
method uses either the digital linear filter method or the logarithmic fast
Fourier transform together with a careful selection of evaluation points and
interpolation. In frequency-to-time domain tests this methodology requires
typically 15 to 20 frequencies to cover a wide range of offsets.
%
The gridding should be frequency- or time-dependent, which is accomplished by
making it a function of skin depth. Optimising for the least number of required
cells should be combined with optimising for computational speed. Looking
carefully at these points results in much smaller computation times with
speedup factors of ten or more over previous methods. A computation in one
domain followed by transformation can therefore be an alternative to
computation in the other domain domain if the required evaluation points and
the corresponding grids are carefully chosen.
%
\end{summary}

@@

The controlled-source electromagnetic (CSEM) method is one of the common
non-seismic tools in exploration geophysics, not only in hydrocarbon
exploration \citep{GEO.10.Constable}, but also in the search for sulfides
\citep{GRL.19.Gehrmann}, water \citep{GEO.05.Pedersen}, geothermal sources
\citep{WGC.15.Girard}, or for geological purposes \citep{NAT.19.Johanson}.
While current sources with a few frequencies are used in the deep marine
environment, transient measurements are more common in the shallow marine
environment and on land \citep[e.g., ][]{GEO.07.Ziolkowski, SEG.07.Andreis,
SEG.07.Avdeeva}. One of the main reasons is the dominance of the airwave in
shallow marine and terrestrial measurements, which can be better separated in
the time domain. CSEM is usually divided into frequency- and time-domain
methods, depending on whether the source signal is a continuous waveform, such
as a sine, or a finite waveform, such as a pseudo-random binary sequence
(PRBS). A numerical comparison of the two methods is given by
\cite{GP.13.Conell}. Acquired CSEM data are subsequently often analysed
(modelled and inverted for) in their respective domain, either the frequency or
time domain. Modellers of layered media usually exploit the horizontal
shift-invariance by computing the responses in the wavenumber-frequency domain
followed by a 2D inverse spatial Fourier transform, also called Hankel
transform, to the space-frequency domain, and a regular inverse Fourier
transform if time-domain data are required \citep[e.g., ][]{GEO.15.Hunziker}.
CSEM codes for 2D and 3D computations, on the other hand, often compute their
responses directly in the required domain, either frequency or time.

Electromagnetic methods in geophysics span a wide range of model scenarios and
acquisition layouts, each with its own modelling-related implications. A
well-known example is the aforementioned land or shallow marine case versus the
deep marine case, where in the former case a dominating airwave has to be
accurately modelled, whereas it can be completely ignored in the latter case.
Recent areas of particular interest in time-domain modelling that pose
numerical challenges are, for instance, simulating the fields through
steel-cased wells \citep{CAG.19.Heagy} or the effects of induced polarization
\citep{EXP.20.Kang}. The former is challenging because of the high conductivity
contrasts requiring very detailed meshing, the latter is challenging for
time-domain modelling because the models are frequency dependent.

Under certain conditions it can be a viable alternative to model time-domain
data with a frequency-domain code, as shown by \cite{GEO.08.Mulder}. They
compared the computational complexity of different time-domain methods with the
computational complexity of frequency-domain computation followed by a Fourier
transform: explicit time-stepping schemes (Du Fort-Frankel method as used in,
e.g., \cite{GEO.04.Commer, GEO.07.Maao}) and implicit schemes \citep[e.g.,
][]{GEO.04.Haber,GEO.10.Um} have a complexity of $O(n^4)$, matrix exponentials
and Lanczos reduction schemes \cite[e.g., ][]{RS.94.Druskin} have a complexity
of $O(n^4\sqrt{\log n})$, and Fourier transform of frequency-domain solutions
have a complexity $O(n_f n^3)$; $n$ is the number of grid points in each
direction, and $n_f$ the number of frequencies. Their conclusion was that the
Fourier transform method can be favourable if the number of required
frequencies is small relative to the number of grid points in each coordinate.
The conditions for fast computation of time-domain data with a frequency-domain
code are: a sufficiently powerful solver, appropriate frequency selection and
interpolation, and an automated gridding, for which they used the
multi-frequency CSEM approach presented by \cite{GEO.07.Plessix}. We build upon
these results but improve the run time from hours to minutes. The main reasons
for this significant speed-up are the further reduction of required frequencies
by introducing lower and upper thresholds of numerically important frequencies,
an adaptive, frequency-dependent gridding scheme that minimizes the required
cells in each dimension, and a logarithmic Fourier transform such as digital
linear filters \citep[DLF, ][]{GP.71.Ghosh} or the logarithmic fast Fourier
transform \citep[FFTLog, ][]{RAS.00.Hamilton} to go from the frequency to the
time domain. The latter makes it also possible to only use the imaginary part
of the frequency-domain response, which has advantages when it comes to
interpolation.

In the next section, we briefly review the methodology as introduced by
\cite{GEO.07.Plessix} and \cite{GEO.08.Mulder} and highlight their advantages
and shortcomings. This is followed by an outline of the methodology, our
changes to the Fourier transform and the adaptive gridding. Finally, we
demonstrate the efficiency of the approach with some numerical results. These
compromise of a homogeneous space, a layered model, a model which includes
dispersive media (induced polarization), and a three-dimensional model.

@@

Being able to model CSEM data both in the frequency domain and in the time
domain can be desirable, as both domains have advantages and disadvantages. One
way to achieve this is to implement Maxwell's equations in both domains as is
done, for instance, in \texttt{SimPEG} \citep{CAG.15.Cockett} or in
\texttt{custEM} \citep{GJI.21.Rochlitz}. Another approach is to have Maxwell's
equations only implemented in one domain, and use Fourier transforms to go to
the other. However, this approach can be costly, as many frequencies over a
wide range are required to go from the frequency domain to the time domain, or
many times over a wide range for the opposite direction. We present a
methodology which significantly reduces the amount and range of the required
frequencies through a combination of extrapolation, interpolation, and setting
the responses for high frequencies to zero. This translates into a significant
reduction in computation time. The number of required frequencies is not the
only important point for reducing computation time. Another important aspect is
the computation grid, as the required grids for low frequencies (in our case in
the order of 0.001\,Hz) and high frequencies (around 100\,Hz) are hugely
different. Low frequencies can be computed on a coarser grid, but they require
a much larger domain in order to avoid boundary effects. High frequencies, on
the other hand, require denser gridding, but they are much more limited in
reach. An adaptive gridding scheme is therefore required, which is naturally
based on the skin depth, the distance after which the amplitude of the
electromagnetic field has decayed by $1/e\approx 37\,\%$. The skin depth
$\delta$ is a function of conductivity and frequency, and for the diffusive
approximation of Maxwell's equations in an isotropic, homogeneous medium is
given by \citep[e.g.,][]{B.SEG.88.Ward}
%
\begin{equation}
  \delta = \sqrt{\frac{2}{\omega\mu\sigma}}
  \ \approx \ 503.3/\sqrt{f\sigma} \, ,
  \label{eq:skindepth}
\end{equation}
%
where $\sigma$ is conductivity (S/m), $\omega=2\pi f$ is angular frequency of
frequency $f$ (Hz), and $\mu$ is magnetic permeability (H/m). The approximation
is obtained by using the free-space value of magnetic permeability,
$\mu_0=4\pi\times10^{-7}\,$H/m.

We build our approach upon \cite{GEO.07.Plessix}, who presented such an
adaptive gridding for multi-frequency (and multi-source) CSEM modelling. They
define the minimum cell width $\Delta_\mathrm{min}$ as a fraction of the
minimum skin depth $\delta_\mathrm{min}$, where $\delta_\mathrm{min}$ should be
2--3 times bigger than $\Delta_\mathrm{min}$. The cells have to be smallest
around the source; in the marine case, the minimum skin depth is therefore
defined by the conductivity of seawater. However, this can yield quite large
cells for low frequencies, so special care has to be taken around the source by
defining a maximum allowed $\Delta_\mathrm{min}$. The grid dimension, on the
other hand, is defined as a function of skin depth for the average conductivity
of the background, $\delta_\mathrm{ave}$. They use four times
$\delta_\mathrm{ave}$ for the $x$-, $y$-, and downward $z$-directions, and a
fixed 50\,km for the upward $z$-direction to account for the airwave. To reduce
the number of cells, it is desirable to introduce stretching, at least in the
buffer zone outside of the area where source and receivers are located.

@@

We present improvements to both the adaptive gridding and the transform from
frequency domain to time domain using the finite-integration technique
\citep{AEU.77.Weiland}, which makes time-domain modelling with a
frequency-domain code viable in comparison with time-domain codes. It is
important to note that our ideas can be used with any solver and is not bound
to one or another frequency- or time-domain solver. The ideas for the frequency
selection can be applied to time selection, and time-dependent, adaptive
gridding exists as well \citep[e.g., ][]{RS.06.Commer}. Our recommendations for
speeding up the Fourier transform for expensive kernels are independent of
spatial complexity, as diffusive EM responses are smooth functions of both
frequency and time. However, the intrinsic requirement is an accurate solver
over a sufficient wide range of frequencies or times.

@@

The requirement for any transform is a robust solver of sufficient accuracy
over a wide range of frequencies. We use for the numerical computations the
open-source (Apache License 2.0) codes \empymod \citep{GEO.17.Werthmuller} and
\emg3d \citep{JOSS.19.Werthmuller}. The former computes semi-analytical
responses of layered models, the latter is a 3D multigrid solver based on
\cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov subspace
solvers or as a solver on its own. The multigrid approach works fine for the
diffusive approximation of Maxwell's equations, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which known improvements such as line-relaxation and semicoarsening
\citep{ECCFD.06.Jonsthovel} are implemented, with a non-standard Cholesky
decomposition to speed up the computation \citep{GEO.08.Mulder}. One of the big
advantages of the multigrid method is that it scales linearly (optimal) with
the grid size in both CPU and RAM usage \citep{B.Springer.20.Mulder}. This
makes it feasible to run even big models on standard computers, without the
need for big clusters. All examples in this article are run on a laptop with an
i7-6600U\,CPU\,@\,2.6\,GHz (x4) and 16\,GB of memory, using Ubuntu 20.04 and
Python 3.8. They were run using a single thread, but parallelization over
frequencies is straight forward, as the computations are independent.

Note that while we use these two algorithms for the numerical examples, any
solver that provides sufficiently accurate frequency-domain responses could be
used.

@@

An important factor in terms of speed and accuracy for time-domain responses
obtained from frequency-domain computations is the selection of the required
frequencies. The fewer frequencies required, the quicker we obtain the
time-domain result. This is true in general but applies in particular to
expensive kernels such as is the case in 3D modelling, where the computation of
the response of a single frequency takes much longer than the transformation
itself. The same applies for the reverse operation, the fewer times required,
the quicker we obtain frequency-domain results. Even though we limit our
analysis here to frequency-to-time transformations, most of the arguments are
reversible and applicable to time-to-frequency transformations. We decided to
use a regular spacing of frequencies on a log-scale, rather than an adaptive
scheme. This approach is favourable if a wide range of offsets is needed, as
the required frequencies change with offset and an adaptive frequency selection
is therefore often tailored to a single offset or a limited range of offsets.
Also, it allows for straightforward parallelization over frequencies, which is
not completely possible with an adaptive scheme.

@@

Any Fourier transform needs a certain range of frequencies to obtain
time-domain results. Our approach is not to minimize the required frequencies
for the transform, but to minimize the number of frequencies for which we
actually have to compute responses. We do this in three zones: above an upper
threshold $f_\mr{max}$, we set the frequency-domain response to zero. Below a
lower threshold $f_\mr{min}$, we extrapolate the frequency-domain response. As
we know that the imaginary response goes towards zero for zero frequency, we
can turn it into an interpolation, for which we use PCHIP. And in-between the
thresholds, we use cubic spline interpolation in order to only have to actually
compute a few distinct frequencies per decade ($n_\mr{dec}^f$). Using this
approach, we found that computing the responses for 15--25 frequencies is
usually good enough to obtain time-domain responses for a large range of
offsets; even though the actual transform might need up to hundreds of
frequencies.

The required frequencies depend naturally on the conductivities and on the
chosen acquisition type and layout. A trial-and-error approach with a 3D code
is very time-intensive. However, a simplified, layered model for the required
survey setup and a fast 1D modeller makes it possible to estimate these
parameters easily. An example of this is shown in Figure~\ref{fig:FFTLog-log}.
%
\begin{figure}
  \centering
  \includegraphics[width=\cwidth]{01-FFTLog-log}
  \caption{Example frequency selection for a simple layered model (the shown
    model parameters are described in the text). This example shows the impulse
    response at an offset of 5\,km, for which it uses FFTLog from 0.001\,Hz to
    10\,Hz with five frequencies per decade.}
  \label{fig:FFTLog-log}
\end{figure}
%
It shows the responses for a marine scenario with 1\,km water depth of
resistivity $\rho=0.3\,\ohmm$ ($\rho = \sigma^{-1}$) and a 100\,m thick target
of 100\,\ohmm at 1\,km below the seafloor in a background of 1\,\ohmm. The
source is 50\,m above the seafloor, the receiver is on the seafloor, and the
response is the inline $x$-directed $E$-field. The left subplot shows the
imaginary part in the frequency domain and the right plot the corresponding
impulse response in the time domain. The red lines are the semi-analytical
responses. The blue circles indicate the actually computed responses and the
black dots the frequencies which are interpolated or set to zero. The resulting
time-domain response has a relative error of less than 1\,\% everywhere except
for very early times.
